On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2020-08-04 manuscripts/thesis/introduction.tex
update

\begin{itemize}
\item Background to the topic.
\item Brief review of current knowledge.
\item Indicate the gap in knowledge. State the aim of the research and how it fits into the gap.
\item Can include an outline of what follows.
\end{itemize}

@@

Sparse coding is a common modern approach to pattern analysis in signal processing whereby each of $N$ observed $n$-dimensional signal samples is viewed as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
%\IEEEPARstart{A}{}common modern approach to pattern analysis in signal processing is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}).

@@

While several algorithms have been recently proposed to provably recover unique dictionaries under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To our knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data has yet to appear in the literature.

@@

More precisely, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. We first consider the following formulation of sparse coding.
%The first mathematical problem we consider is the following.

@@

Our development of Thm.~\ref{DeterministicUniquenessTheorem} is general enough to provide some uniqueness and stability even when generating $\mathbf{A}$ do not fully satisfy (\ref{SparkCondition}) and recovery dictionaries $\mathbf{B}$ have more columns than $\mathbf{A}$.  Moreover, the approach incorporates a combinatorial theory for designing generating codes that should be of independent interest. We also give brief arguments adapting our results to dictionaries and codes drawn from probability distributions (Cor.~\ref{ProbabilisticCor}). The technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Sec.~\ref{DUT}, following some necessary definitions and a fact in combinatorial matrix analysis (Lem.~\ref{MainLemma}; proven in the Appendix). Finally, we discuss in Sec.~\ref{Discussion} applications of our mathematical observations as well as open questions. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

>>>>

%\begin{itemize}
%\item Background to the topic.
%\item Brief review of current knowledge.
%\item Indicate the gap in knowledge. State the aim of the research and how it fits into the gap.
%\item Can include an outline of what follows.
%\end{itemize}

@@

Sparse coding is a common modern signal processing technique that views each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
%\IEEEPARstart{A}{}common modern approach to pattern analysis in signal processing is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}).

@@

While several algorithms have recently been proposed to provably recover unique dictionaries under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To our knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data has yet to appear in the literature.

@@

More precisely, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. We first consider the following decidable\footnote{Note that Prob.~\ref{InverseProblem} is decidable for rational inputs $\mathbf{z}_i$ \cite{chrishillar} since the statement that it has a solution can be expressed as a logical sentence in the theory of algebraically closed fields, and this theory has quantifier elimination \cite{basu2006algorithms}.} formulation of sparse coding.
%The first mathematical problem we consider is the following.

@@

Our development of Thm.~\ref{DeterministicUniquenessTheorem} is general enough to provide some uniqueness and stability even when generating $\mathbf{A}$ do not fully satisfy (\ref{SparkCondition}) and recovery dictionaries $\mathbf{B}$ have more columns than $\mathbf{A}$.  Moreover, the approach incorporates a theory of combinatorial designs for the sparse supports of generating codes $\mathbf{x}_i$ that should be of independent interest. We also give brief arguments adapting our results to dictionaries and codes drawn from probability distributions (Cor.~\ref{ProbabilisticCor}). The technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Sec.~\ref{DUT}, following some necessary definitions and a fact in combinatorial matrix analysis (Lem.~\ref{MainLemma}; proven in the Appendix). We discuss these results and their applications in depth in Sec.~\ref{Discussion} and outline open questions and directions for future research in Sec.~\ref{FutureDirections}. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

### 2020-08-04 manuscripts/thesis/results.tex
update

Consequently, if $k$ is held fixed or if the size of the support set of reconstructing codes is polynomial in $\overline m$ and $k$, then a practical (polynomial) amount of data suffices to identify the dictionary.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, Tillmann15}. Even so, in the next section we discuss how probabilistic guarantees can in fact be made for any number of available samples.

>>>>

Consequently, if $k$ is held fixed or if the size of the support set of reconstructing codes is polynomial in $\overline m$ and $k$, then a practical (polynomial) amount of data suffices to identify the dictionary.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}. Even so, in the next section we discuss how probabilistic guarantees can in fact be made for any number of available samples.

### 2020-08-04 manuscripts/thesis/reviewers.tex
update

Given a dictionary A and codes x-i that solve Problem 1 (e.g., learned by any algorithm), check that that they satisfy the assumptions of Thm. 1:
   - List the support sets of the x-i.
   - Discard those for which there are (k-1)(m /choose k) or fewer x-i with that support.
   - Discard those for which the supported x-i are not in general linear position. (This should also exclude any supports less than k in size.)
   - From the support sets that remain, list all SIP-satisfying hypergraphs that are subsets of this set.
   - Discard those for which L-(2H)(A) = 0.
   - For each of the remaining hypergraphs, determine the constant C-1 from A and the x-i (or for every subset of the x-i of size (k-1)(m /choose k)).
   - Check that the derived upper bound on /varepsilon exceeds that of the original problem. If yes, the solution is "unique".

@@

@@

NOTE: Structured dictionaries? Just derive the bounds specific to them. Could be efficiently computable in specific cases.

REVIEWER

As l0 norm is intractable, does all the properties still hold for l1 norm minimization? Can we solve it efficiently (polynomial time) to global optimality? Similar issue for other works, such as
https://arxiv.org/pdf/1807.05595.pdf

RESPONSE:
We make no claims in this paper about l1-norm minimization, though we are hopeful a clever reader may find some way to build off our results and derive related guarantees for this continuous relaxation of the sparse coding problem.

>>>>

Given a dictionary A and codes  that solve Problem 1 (e.g., learned by any algorithm), check that that they satisfy the assumptions of Thm. 1:
   - List the support sets of the $\mathbf{x_i}$.
   - Discard those for which there are $(k-1){m \choose k}$ or fewer \mathbf{x_i} with that support.
   - Discard those for which the supported $\mathbf{x_i$} are not in general linear position. (This should also exclude any supports less than k in size.)
   - From the support sets that remain, list all SIP-satisfying hypergraphs that are subsets of this set.
   - Discard those for which $L_{2\mathcal{H}}(\mathbf{A}) = 0$.
   - For each of the remaining hypergraphs, determine the constant $C_1$ from \mathbf{A} and the $\mathbf{x_i}$ (or for every subset of the  $\mathbf{x_i}$ of size $(k-1){m \choose k}$).
   - Check that the derived upper bound on $\varepsilon$ exceeds that of the original problem. If yes, the solution is "unique".

@@

!HERE!

@@

### 2020-08-04 manuscripts/thesis/thesis.tex
update

% *** ALIGNMENT STUFF***
\usepackage{array}
\usepackage{bm}
\usepackage{rotating} % provides sidewaystable and sidewaysfigure

@@

\begin{dedication}
\null\vfil
%\begin{center}
Dedication
%I want to dedicate this to my dad. At the same time, I don't want to, because the results on display here can't be matched to the time with him I traded to produce them, and it feels as though dedicating this work to him would be measuring his life up against the theorems contained here-in. I knew very well this work wasn't so important; these theorems, while fundamental, are just something people might cite to feel better about whatever sketchy sparse coding analysis they were already going to do, anyway. And yet, despite total freedom offered by the Helen Wills Neuroscience Institute, I worried the brain was so dauntingly complex that I should feel lucky to have at last been offered a well-defined relevant problem to work on, particularly one I felt even remotely equipped to solve myself, so as to feel deserving of a Ph.D. from UC Berkeley. The fact is that whatever the eventual effect of this work on the scientific ecosystem, these truths could have waited, arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited for death. I couldn't really fathom what it actually meant for him to die, to never see or talk to or hug him again, while the idea of putting original work out there, of crafting my own piece of the overwhelming puzzle that is neuroscience, was tantalizing. Let's say, then, that this dedication is the first bit of a life's work, real work void of any prestigious position or degree and the associated self-doubt and pressure, put toward passionate id	eas, solving problems I care about and that actually matter, loving and caring for others or, as my dad would say, ``being a mensch". May this dedication serve to hold myself publicly accountable to that. Also I love you, mom. I dedicate and don't dedicate this and everything to come to you, too.
\vspace{12pt}
%\end{center}
\vfil\null
\end{dedication}

@@

\begin{acknowledgements}
I would like to thank Chris Hillar for laying down the groundwork and vetting all my proofs. He gave me a lot of his time for little if anything in return. Fritz Sommer for taking on the role of my advisor. The Berkeley Student Cooperative for a really awesome and transformative time. CZ or Die. 
We thank Darren Rhea for early thoughts, and Ian Morris for posting \eqref{eqdim} online.

\end{acknowledgements}

@@

% (Optional)
 \part{Main Text}

\include{introduction}
\include{results}
\include{proofs}
\include{discussion}
\include{conclusions}

\include{appendix} 
%\include{conclusions}

>>>>

% *** ALIGNMENT STUFF***
\usepackage{array}
\usepackage{bm}
\usepackage{rotating} % provides sidewaystable and sidewaysfigure
\usepackage{url}

@@

%\begin{dedication}
%\null\vfil
%\begin{center}
%I want to dedicate this to my dad. At the same time, I don't want to, because the results on display here can't be matched to the time with him I traded to produce them, and it feels as though dedicating this work to him would be measuring his life up against them. These theorems, while fundamental, are just something people might cite to feel better about the sparse coding analyses they were already going to do, anyway. And yet, despite total freedom offered by the Helen Wills Neuroscience Institute, I worried the brain was so dauntingly complex that I should feel lucky to have at last been offered a well-defined relevant problem to work on, particularly one I felt even remotely equipped to solve myself, so as to feel deserving of a Ph.D. from UC Berkeley. The fact is that whatever the eventual effect of this work on the scientific ecosystem, these truths could have waited, arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited for death. I couldn't really fathom what it actually meant for him to die, to never see or talk to or hug him again, while the idea of putting original work out there, of crafting my own piece of the overwhelming puzzle that is neuroscience, was tantalizing. Let's say, then, that this dedication is the first bit of a life's work, real work void of any prestigious position or degree and the associated self-doubt and pressure, put toward passionate id	eas, solving problems I care about and that actually matter, loving and caring for others or, as my dad would say, ``being a mensch". May this dedication serve to hold myself publicly accountable to that. Also I love you, mom. I dedicate and don't dedicate this and everything to come to you, too.
%\vspace{12pt}
%\end{center}
%\vfil\null
%\end{dedication}

@@

%\begin{acknowledgements}
%I would like to thank: Chris Hillar for laying down the groundwork and vetting all of my proofs many times over. He lent me a lot of his time and expertise for little in return. Fritz Sommer for advising me and pointing me toward a tractable problem to solve. The Berkeley Student Cooperative for the amazing experience living co-operatively in Berkeley. CZ or die.
%\end{acknowledgements}

@@

\section{Foreword}
This thesis presents results published Dec.~2019 in a paper of the same name in \emph{IEEE Transactions on Signal Processing}, largely verbatim. Readers should refer to the Discussion and Future Directions as well as Figures \ref{probpigeon} and \ref{reasonableC2} for additional insights not offered in the paper.

% (Optional)
 %\part{Main Text}

\include{introduction}
\include{results}
\include{proofs}
\include{discussion}
%\include{reviewers}
\include{appendix}
